Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-12-01 response_to_referees.tex

\refa{1.1. Mutations in influenza are highly nested, with a substantial
hitch-hiking, and no attempt is made to disentangle such dependencies when counting the mutations.}

\response{This is a good suggestion, and we agree with reviewer 1 that our original manuscript lacked a method to disentangle nested mutations. \\
We added a new section in the Supplementary Material where we attempt to cluster together trajectories of mutations that partly appear on the same strains. 
Trajectories corresponding to mutations always or often appearing on the same strains are then counted as one "effective" trajectory. 
However, this new way of counting mutations does not significantly change our results, and as a consequence we left the figures of main text unchanged.}

\refa{1.2 Despite the more general formulation in the beginning, this
approach makes use only of the last time point in the trajectory,
rather than the full trace (contrary to the work of Illingworth\&
Mustonen). I find the observed neutral-like statistics not surprising
for such a limited data input, which doesn't capture past frequency dynamics. Therefore, these conclusions should be
revisited and benchmarked against the more general implementation of Illingworth\&Mustonen-like approach.}

\responsedraft{
	We thank reviewer 1 for pointing out the work of Illingworth\&Mustonen to us, which we were unaware of. However, we believe that the method used in their article is not applicable as is in our case. To explain our thinking, we first briefly summarize the approach taken in Illingworth\&Mustonen: the authors fit two kind of models to fully known frequency trajectories: 
	\begin{itemize}
		\item An "unlinked" model, where the fitness effect of a mutation fully determines its ultimate fixation or loss. The only trajectories that the model can produce are upward or downward sweeps, for beneficial or deleterious mutations respectively.
		\item A "linked" model, in which the fate of a mutation is determined both by its own fitness and by the fitness of other mutations present on the genomes it appears on. This model fits actual trajectories much more accurately, which is one of the main findings of the paper.
	\end{itemize} 
	Interestingly, both models depend on the same parameters that have to be fitted, namely the fitness effects of individual mutations. Linkage between mutations in the second model is introduced by using the measured frequencies of appearance of two mutations on the same genome. \\
	\pierre{What I'm trying to say below
	\begin{itemize}
		\item Illingworth\&Mustonen don't do prediction, so there is no clear way to apply their method to our article.
		\item Fitting the initial part of the trajectory using their models doesn't make much sense. 
		\item We want to stay model free: it's possible to find signs of selection without fitting (cf. H1N1/simulated data)
		\item Even if we're doing something very simple, it's surprising to see neutral-like behaviour. 
	\end{itemize}}
	We first want to point out that this approach is fundamentally different to the one of our article. Since Illingworth\&Mustonen fit models to fully known trajectories, these models cannot meaningfully be used for prediction purposes. Indeed, fitness effects of individual mutations are fitted with knowledge of the fixation or loss of the given mutation, and of the frequency of other mutations in the case of the second model. This strongly contrasts with our approach which is to find patterns of predictability of the future behaviour of frequency trajectories using past data only: \emph{e.g.} if a frequency trajectory has risen from $0$ to a frequency $f>0$, what can be said of its future shape or of its fixation? Thus, methods in Illingworth\&Mustonen are designed to answer a different question than the one we are asking in our work. \\
	A way of reconciling the two approaches would be for us to fit the initial part of a trajectory with either model, and to use this fit to predict its future behaviour. However, we feel that this idea is not very relevant to our article for two reasons. First, neither model proposed by Illingworth\&Mustonen is really practical in our setting. The "unlinked" model only generates sweep-like trajectories, which our results show to be not representative of the "typical" trajectory (in good agreement with Illingworth\&Mustonen). The "linked" model can fit more complex trajectories, \emph{e.g.} ones that rise rapidly but ultimately die. However, it crucially depends on the frequencies of joint appearance of pairs mutations on genomes to introduce linkage. These frequencies vary through time, allowing the \emph{effective} fitness of a mutation to vary, and a trajectory to rise and then fall. In our case, these frequencies would only be known for the past, making impossible any prediction of the model. \\
	The second reason is that we believe that our model-free approach has intrinsic benefits. [Something saying that we'd prefer not to start fitting models to  data, because it's not the point. But I'm not sure how to formulate this.] \\
	Finally, we are aware that using only the last time point of a rising trajectory gives only limited opportunity for predicting its future. However, observing apparently-neutral statistics in the case of A/H3N2 is still surprising to us. The case of simulated populations shows that clear signs of selection can be observed by looking at basic features of trajectories (\emph{e.g.} figure S20 of the supplementary material). This is consistent with the theory of evolving population: if the only available information about a mutation is that it has risen in frequency starting from $0$, it has a higher chance on being beneficial than deleterious, and its fixation probability should be higher than what it would be in a neutral scenario. Hence, even with our very simple approach, we observe that predictability of A/H3N2 qualitatively differs from what models of adapting populations would suggest. \\}

@@

\refb{2. Fig 2A and 2B look quite different to me. In the text it appears to me as if these are very similar to the authors. What are the characteristics of the AAs that fix in H1N1? }

\refb{3. Question: what do these results mean for vaccines? How to decide which strains to use for vaccines? This may be known to those who work on Influenza, but for a relative outsider it is not clear. }

\refb{4. Question: at what point is it expected that the vaccine itself will influence frequency trajectories? See Wen et al Biorxiv 2020}
\responsedraft{We should mention Wen \emph{et. al.} MDPI-Viruses 2018: \emph{Estimating Vaccine-Driven Selection in Seasonal Influenza}. 
The answer is that yes it's expected that vaccine influence frequency trajectories to some extent, but it is very hard to measure in practice. 
The above paper attempted to measure this, and did not find clear-cut results.}

\refb{Fig 1B is very hard to read. Maybe it should get more space (fig 1C could work with less space).}
\responsedraft{We have reworked this figure to allow more space for panel B.}

### 2020-12-04 article.tex
Refer to sec3 of SI

%Frequency trajectories allow a simple formulation of the above-stated questions. The short term prediction amounts to predicting the curve $f(t)$ starting from a given date $t_0$. On the other hand, the long term prediction amounts to predicting how long a trajectory will remain active, and whether it will fix or disappear when it finishes. 
In the rest of this work, we will focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$. 
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one. 
If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants. 
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fixation. 
The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments. 
By considering the ensemble of novel variants that are rising in frequency, we effectively average over backgrounds, obtaining a set of mutations that we expect to be beneficial on average.
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.

@@

However, future distributions for $\Delta t >0$ do not seem to follow a pattern compatible with selective sweeps. 
The thick black line in Figure~\ref{fig:trajectories}B shows the average frequency of all trajectories. 
This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$ in the case of A/H3N2, and slightly increasing for A/H1N1pdm (see supplement).
Hence, the fact that this average rose for $t<0$ gives little information for $t>0$, and is due to the conditions by which these trajectories were selected.
This shows that sweep-like trajectories rising steadily from frequency $0$ to $1$ are not common enough to dominate the average trajectory.

@@

We then examined the probability of mutations to fix in the population as a function of the frequency at which they are seen. 
For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time. 
The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} show $P_{fix}(f)$ as a function of $f$ for NA and HA. 
For both proteins, the probability of fixation of a new mutation at frequency $f$ is close to $f$ itself, that is $P_{fix}(f)\simeq f$. 
This result is exactly what is expected in a population evolving in the \emph{absence} of selection. 
% Within the diffusion approximation of the Wright-Fisher model~\cite{kimura_diffusion_1964}, the fixation probability of a mutation with fitness effect $s$ at frequency $f$ is
% \begin{equation}
%     P_{fix}(f) = \frac{1-e^{-2sfN}}{1-e^{-2sN}} \approx
%     \begin{cases}
%     1-e^{-2sfN} & sfN \gg 1 \\
%     f & |sN| \ll 1 \\
%     e^{-2s(1-f)N} & -sN \gg 1  
%     \end{cases}
% \end{equation}
% The result for the neutral case is intuitive:
A mutation or trait appearing at frequency $f$ is shared by $f\cdot N$ individuals, and the probability for one of them to become the ancestor of all the future population is $f\cdot N/N=f$. 
Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. 
Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation, while those in A/H1N1pdm show only small deviations from it. 
In both cases, the probability of fixation seems to be mainly dictated by the current frequency $f$ at which the mutation is observed.

@@

\subsection{Correcting for nested mutations} % (fold)
\label{sub:correcting_for_nested_mutations}

@@

\begin{figure}
	\centering
	\includegraphics[scale=0.5]{Rev_figures/clusters_examples.png}
	\caption{\textbf{Top}: Left-axis: number $n$ of clusters as a function of the threshold distance $d^\star$. The total number of trajectories is shown as a flat orange line. Right-axis: average size of clusters as a function of $d^\star$. \textbf{Bottom}: Examples of clusters for four values of $d^\star$. The four clusters displayed are the ones to which mutation HA1:33R belongs to. }
	\label{fig:cluster_examples}
\end{figure}

>>>>

Since segments in influenza do not recombine, trajectories defined as above are sometimes heavily correlated with each other. 
Indeed, the trajectories corresponding to two mutations that partly appear on the same background will tend to show a similar behavior. 
In the extreme case were the two mutations always appear on the same strains, their two trajectories will be identical. 
This ``nested'' characted or frequency trajectories may introduce biases when performing statistical analysis. 
In section \ref{sub:correcting_for_nested_trajectories} of the SM, we attempt to correct for nested trajectories by clustering them based on the strains that compose them. 
However, this does not significantly change the results presented below.

In the rest of this work, we will focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$. 
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one. 
If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants. 
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fixation. 
The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments. 
By considering the ensemble of novel variants that are rising in frequency, we effectively average over backgrounds, obtaining a set of mutations that we expect to be beneficial on average.
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.

@@

However, future distributions for $\Delta t >0$ do not seem to follow a pattern compatible with selective sweeps. 
The thick black line in Figure~\ref{fig:trajectories}B shows the average frequency of all trajectories. 
This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$ in the case of A/H3N2, and slightly increasing for A/H1N1pdm (see supplement).
Hence, the fact that this average rose for $t<0$ gives little information for $t>0$, and is due to the conditions by which these trajectories were selected.
	This shows that sweep-like trajectories rising steadily from frequency $0$ to $1$ are not common enough to dominate the average trajectory.  \pierre{This result is consistent with\citep{illingworth_components_2012}. They show that sweep-like trajectories are very bad fits to actual trajectories.}

@@

We then examined the probability of mutations to fix in the population as a function of the frequency at which they are seen. 
For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time. 
The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} show $P_{fix}(f)$ as a function of $f$ for NA and HA. 
For both proteins, the probability of fixation of a new mutation at frequency $f$ is close to $f$ itself, that is $P_{fix}(f)\simeq f$. 
This result is exactly what is expected in a population evolving in the \emph{absence} of selection. 
% Within the diffusion approximation of the Wright-Fisher model~\cite{kimura_diffusion_1964}, the fixation probability of a mutation with fitness effect $s$ at frequency $f$ is
% \begin{equation}
%     P_{fix}(f) = \frac{1-e^{-2sfN}}{1-e^{-2sN}} \approx
%     \begin{cases}
%     1-e^{-2sfN} & sfN \gg 1 \\
%     f & |sN| \ll 1 \\
%     e^{-2s(1-f)N} & -sN \gg 1  
%     \end{cases}
% \end{equation}
% The result for the neutral case is intuitive:
A mutation or trait appearing at frequency $f$ is shared by $f\cdot N$ individuals, and the probability for one of them to become the ancestor of all the future population is $f\cdot N/N=f$. 
Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. 
Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation, while those in A/H1N1pdm show only small deviations from it. \pierre{``limited but clear'' instead of ``small''?}
In both cases, the probability of fixation seems to be mainly dictated by the current frequency $f$ at which the mutation is observed.

@@

\subsection{Correcting for nested trajectories} % (fold)
\label{sub:correcting_for_nested_trajectories}

@@

\begin{figure}
	\centering
	\includegraphics[scale=0.5]{Rev_figures/clusters_examples.png}
	\caption{\textbf{Top}: Left-axis: number $n$ of clusters as a function of the threshold distance $d^\star$. The total number of trajectories is shown as a flat orange line. Right-axis: average size of clusters as a function of $d^\star$. \textbf{Bottom}: Examples of clusters for four values of $d^\star$. The four clusters displayed are the ones to which the trajectory of mutation HA1:33R belongs. }
	\label{fig:cluster_examples}
\end{figure}
